The Weakly Nonlinear Magnetorotational Instability in a Global, Cylindrical Taylor Couette Flow

### 2016-09-11 paper/widegap_mri_paper.tex
updated nsf grant no

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system (\citei{Parker:2013hy}). Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (\citei{Johansen:2009uj}, \citei{Kunz:2013}). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the application of this model system to realistic astrophysical disks. Our model is most directly relevant to TC flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

@@

\begin{itemize}
\item Our theory may be applied to a specific experimental apparatus to model the predicted saturated state. One can then ask whether GLE dynamics should be detectable, especially over endcap-driven flows.
\item The saturation properties of different rotation profiles may be compared by direct application of the theory developed here.
\item Vertical stratification may be added to the base state, constructing a more realistic model of global vertical disk structure. 
\item Other nonideal MHD effects such as the Hall effect and ambipolar diffusion are straightforward addions to this model, and are of particular interest for understanding protoplanetary disks. 
\item The background magnetic field geometry may be generalized to include radial variation, another feature relevant to astrophysical disks.
\item The radial boundary conditions considered here may be expanded to mimic astrophysical disks rather than TC devices.
\item Our theory can be compared to simulations in both the weakly and strongly nonlinear regimes: both in the pattern selection at saturation and the $\Pm^{0.777}$ scaling can be directly tested. 
\end{itemize}

@@

\section{Acknowledgments}
S.E.C. was supported by a National Science Foundation Graduate Research Fellowship under grant No. DGE-11-44155. J.S.O. acknowledges support from NASA grant 15-LWS15\_2-0063. The authors thank Mordecai Mac Low, Jeremy Goodman, John Krommes, Geoff Vasil, and Ellen Zweibel for useful discussion.

@@

>>>>

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system \citep{Parker:2013hy, Parker:inpress}. Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (\citei{Johansen:2009uj}, \citei{Kunz:2013}). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the application of this model system to realistic astrophysical disks. Our model is most directly relevant to TC flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

@@

\begin{itemize}
\item Our theory may be applied to a specific experimental apparatus to model the predicted saturated state. One can then ask whether GLE dynamics should be detectable, especially over endcap-driven flows.
\item The saturation properties of different rotation profiles may be compared by direct application of the theory developed here.
\item Vertical stratification may be added to the base state, constructing a more realistic model of global vertical disk structure. 
\item Other nonideal MHD effects such as the Hall effect and ambipolar diffusion are straightforward additions to this model, and are of particular interest for understanding protoplanetary disks. 
\item The background magnetic field geometry may be generalized to include radial variation, another feature relevant to astrophysical disks.
\item The radial boundary conditions considered here may be expanded to mimic astrophysical disks rather than TC devices.
\item Our theory can be compared to simulations in both the weakly and strongly nonlinear regimes: both the pattern selection at saturation and the $\Pm^{0.777}$ scaling can be directly tested. 
\end{itemize}

@@

\section{Acknowledgments}
S.E.C. was supported by a National Science Foundation Graduate Research Fellowship under grant No. DGE-16-44869. J.S.O. acknowledges support from NASA grant 15-LWS15\_2-0063. The authors thank Mordecai Mac Low, Jeremy Goodman, John Krommes, Geoff Vasil, and Ellen Zweibel for useful discussion.

@@

\bibitem[{Parker \& Krommes(in press)}]{Parker:inpress}
Parker, J.B. \& Krommes, J.A., in press, in Zonal Jets, ed. B. Galperin \& P. Read (Cambridge, U.K.: Cambridge)
